### 2025-06-04 paper/relatedwork.tex
More related work.

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax,radul2023you}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Note that while our work
does not give a formal proof that AD implementation is correct (it is
not clear how to formulate correctness criteria for it), the entire
implementation of AD is about 20 lines of code, so it is easy to
understand and modify.

Another line of work studies high-level principles of AD using
category theory~\cite{ad-theor1, ad-theor2, ad-theor3}.
While this indeed comes with great correctness guarantees due to
some naturality principles, it is not always clear how to implement
this in a way that leads to efficiently executable specifications.  Also, the
entire treatment of index-safe tensors is unclear.

>>>>

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad} and Haskell~\cite{ad-haskell}.
%
Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
%
Most of these approaches are largely similar to ours (although naturally more
complete), as they involve a syntactically driven transformation of the program.

While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Note that while our work
does not give a formal proof that AD implementation is correct (it is
not clear how to formulate correctness criteria for it), the entire
implementation of AD is about 20 lines of code, so it is easy to
understand and modify.

The AD transformations in Dex~\cite{10.1145/3473593} and Jax~\cite{ad-jax} are
in this respect interesting, as they are based on first \emph{linearising} the
function of interest in a given point (essentially a forward mode
transformation), then \emph{transposing} the resulting linear map to obtain a
reverse mode linear map~\cite{radul2023you}. This approach might make be easier
to implement in a fully verified manner, as linearisation and transposition have
well-defined properties and could be verified independently.

Another line of work studies high-level principles of AD using category
theory~\cite{ad-theor1, ad-theor2, ad-theor3}. While this indeed comes with
great correctness guarantees due to some naturality principles, it is not always
clear how to implement this in a way that leads to efficiently executable
specifications, although some efforts are underway~\cite{10.1145/3632878}. Also,
the entire treatment of index-safe tensors is unclear.
